A community-maintained standard library of population genetic models

### 2020-05-05 popSimManu.tex
updated figures and src

For these reasons, we have generated a standardized, community-driven resource
for simulating published demographic models from a number of popular study systems.
This resource, which we call \stdpopsim, makes running
realistic simulations for population genetic analysis a simple matter of
choosing pre-implemented models from a community-maintained catalog.
The \stdpopsim catalog currently contains five species: humans,
\textit{Drosophila melanogaster}, \textit{Arabidopsis thaliana},
\textit{Escherichia coli}, and \textit{Canis familiaris}.
For each species, the catalog contains information on our current understanding of
the physical organization of its genome (e.g., chromosome structure),
one or more inferred genetic maps (except for \textit{E.~coli}),
and default population-level parameters (mutation rate and generation time estimates).
The catalog also contains numerous published demographic models for the first three species.
These available models and parameters are meant to represent the field's current understanding,
and we intend for this resource to evolve as new results become available, and relevant
existing models are added to \stdpopsim by the community.
Through either a command line interface or a simple Python API, users can specify which
organism, genetic map, chromosome, and demographic history they are interested in simulating, and the
simulation output from their chosen model is returned.
In this way, \stdpopsim will lower the barrier to high-quality simulation for exploratory analyses,
enable rigorous evaluation of population genetic software,
and contribute to increased reliability of population genetic inferences.

@@

%ACS: the consortium really isn't the point; let's get right to describing what stdpopsim is
%The first contribution of the PopSim consortium is 
The \stdpopsim library is a
a community-maintained collection of empirical genome data and population genetics simulation models.
Figure \ref{fig:cartoon} shows a graphical representation of the structure of \stdpopsim.
The package centers on a catalog of genomic information and demographic models
for a growing list of species (Fig.~\ref{fig:cartoon}A), 
and software resources to facilitate efficient simulations (Fig.~\ref{fig:cartoon}B-C).

@@

Given the genome data and simulation model descriptions defined within the
library, it is straightforward to run standardized simulations
across a range of organisms. \Stdpopsim has a Python API and a user-friendly
command line interface, allowing users with minimal experience direct access to
state-of-the-art simulations. Simulations are output in the ``tree sequence''
format \citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring}, which
contains complete genealogical information about the simulated samples, is
extremely compact, and can be processed efficiently using the \tskit library
\citep{kelleher2016efficient,kelleher2018efficient}.
The tree sequence format could also be converted
to other formats (e.g., VCF) by the user if desired.
Currently, \stdpopsim uses the  \texttt{msprime} coalescent simulator \citep{kelleher2016efficient}
as the default simulation engine.

Coalescent simulations, while highly efficient, are limited in their ability to model
complex selection scenarios, such as recurrent sweeps and background selection. For these reasons,
we have also implemented the forward-time simulator, \texttt{SLiM} \citep{haller2019tree,haller2019slim},
as an alternative backend engine to \texttt{stdpopsim}, allowing for the simulation of processes that cannot be modeled under
the coalescent. However, as forward-time simulators explicitly model all individuals in a population, simulating large
population sizes can be highly demanding of computational resources. One common practice used to address this challenge is to
scale down the population size used in the simulations (\textit{N}),
scale down times specified in the demographic model by the same factor, and scale up the mutation and recombination
rates and selection coefficients such that the scaled parameters, \textit{N$\mu$}, \textit{Nr},
and \textit{Ns}, remain unchanged \citep[see the \texttt{SLiM} manual and][]{uricchio2014robust}.
Such down-scaled simulations are not completely equivalent to simulating
all individuals in the population, and may lead to subtle differences.
However, since many sequence-based measures of population diversity remain unchanged when rescaling in this fashion,
this practice is effective for many purposes and widely employed.

We validated our implementation of the \texttt{SLiM} engine by comparing
estimates of several population genetic summary statistics for neutral simulations generated by both \texttt{SLiM}
and \texttt{msprime}.
Examples of this validation for different down-scaling factors are shown in
Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}, which indicate that scale factors of up to
$50$ ($Q=50$ in \texttt{SLiM}) produce patterns of both diversity and linkage disequilibrium that are
qualitatively similar to those observed under the coalescent (i.e. \texttt{msprime}), when both are simulated under a recombination map
(Figure \ref{fig:slim_val_map}).
We do observe a slight departure in values of Tajima's $D$ when scaling down by a factor of $50$.
%likely due to the subtle differences produced by down-scaling.
Interestingly, the effects of down-scaling are more prominent and extend to other sequence-based statistics
when data sets are simulated under a uniform recombination rate (Figure \ref{fig:slim_val_nomap}).
Importantly, both comparisons validate the equivalence of \texttt{SLiM} and \texttt{msprime} when no down-scaling is applied,
and they suggest that down-scaling population sizes can significantly reduce running time at a fairly modest cost to accuracy.
% ACS: We might consider deleting the next two sentences.  This feels like a lot on down-scaling, and it's a bit verbose as written.  Maybe 
% better to come back to it in the discussion if we want to say more.
However, since the effects of rescaling are not fully understood, scaling factors should be used as conservatively as practicable.
The framework provided by \texttt{stdpopsim} should enable a deeper inspection of this issue, which we expect will contribute to
best practices used when applying forward-time simulators such as \texttt{SLiM}.

%Ilan: added this short subsection since this paragraph didn't quite fit any of the other subsections
\subsection*{Documentation and reproducibility}

@@

% ACS: there was indeed a lot of redundancy here.  DRS improved it a lot, but I've tried to do a bit more
\subsubsection*{ Single-population demographic models.}
For single-population demographic models, we compared
\MSMC~\citep{schiffels2014inferring}, \smcpp~\citep{terhorst2017robust}, and
\stairwayplot~\citep{liu2015exploring}
 on simulated genomes sampled from a single population,
under several of the demographic models described above.
However, these experiments raise the question of what to use as the ``true" population sizes in the case of multi-population models with migration.
In particular, when simple single-population models are fit to data simulated under such complex models, 
they are not expected to recover the actual simulated population sizes owing to model misspecification.
Instead, we argue that the best one may expect in such a scenario is to infer a model that accurately reflects the
coalescence time distribution of the simulated model.  In turn, the inverses of these coalescence times (expressed in appropriate units) 
correspond to population sizes.
Under a multi-population model, the coalescence time distribution is influenced by migration between the target population and populations not analyzed in inference,
as well as by the ancestral effective population sizes.
We thus analytically computed inverse coalescence
rates in \texttt{msprime} for each simulated model, and used them as benchmarks for the ``true'' effective population sizes.
See the Appendix for a precise definition and description of the inverse coalescence rate computation.

@@

In Figure \ref{fig:IM_popn_human}C we show estimates of population sizes and divergence
time, for each of the inference methods, using samples drawn from African and European populations
simulated under the OutOfAfrica\_3G09 model. Our results highlight many
of the strengths and weaknesses of the different methods.
For instance, the SFS-based approaches with simple IM models do not capture
recent exponential growth in the CEU population, but do consistently recover the
simulated YRI population size history. Moreover, these approaches allow 
migration rates to be estimated (Figure \ref{fig:homsap_mig_rates}), and lead to more accurate inferences
of divergence times. However, these migration rate estimates are somewhat biased.
%ACS: seems unnecessary and redundant
%likely due to model misspecification.
In contrast, \smcpp is much better at capturing the recent exponential
growth in the CEU population, though the inferred population sizes are generally noisier.
In addition, \smcpp consistently underestimates divergence times
because it assumes no migration between populations (Figure \ref{fig:IM_popn_human}C).

@@

\paragraph{Benchmarking population size inference.}
We have illustrated in this paper how \stdpopsim can be used for direct
comparisons of inferential methods on a common set of simulations. Our
benchmarking comparisons have been limited, but nevertheless
reveal some informative features.
For example, at the task of estimating population size histories for simulated human
populations, we find that the sequence-based methods (\MSMC and \smcpp)
perform somewhat better overall---at least for moderate times in the past---than
the site frequency spectrum-based method (\stairwayplot),
which tends to over-estimate the sizes of oscillations
(Figures~\ref{fig:n_t_ragsdale} and~\ref{fig:n_t_gutenkunst}).
In contrast, \stairwayplot outperforms the sequence-based methods
on simulations of \textit{D.~melanogaster} or \textit{A.~thaliana} populations,
in which linkage disequilibrium is reduced (Figures~\ref{fig:n_t_sheehan} \& \ref{fig:n_t_AraTha}).
In simulations of two human populations
(Figure~\ref{fig:IM_popn_human}), \dadi and \fastsimcoal do reasonably well at
reconstructing the simulated YRI history, but struggle with the more complex simulated CEU
history, in large part because the methods assume constant population sizes.
On the other hand, \smcpp does not have the same
restrictions on its inferred history, and as a result does somewhat better
with the CEU history but tends to overcomplicate the simple YRI history.
The results for
the two-population \textit{D.~melanogaster} model (Figure~\ref{fig:two_popn_fly})
are generally similar.
In these comparisons, \fastsimcoal and \dadi perform almost identically, which is expected because they fit the same models to the same summaries of the data, differing only in how they calculate model expectations and optimize parameters.

>>>>

For these reasons, we have generated a standardized, community-driven resource
for simulating published demographic models from a number of popular study systems.
This resource, which we call \stdpopsim, makes running
realistic simulations for population genetic analysis a simple matter of
choosing pre-implemented models from a community-maintained catalog.
The \stdpopsim catalog currently contains five species: humans,
\textit{Drosophila melanogaster}, \textit{Arabidopsis thaliana},
\textit{Escherichia coli}, and \textit{Canis familiaris}.
For each species, the catalog contains information on our current understanding of
the physical organization of its genome (e.g., chromosome structure),
one or more inferred genetic maps (except for \textit{E.~coli}),
and default population-level parameters (mutation rate and generation time estimates).
The catalog also contains numerous published demographic models for the first three species.
These available models and parameters are meant to represent the field's current understanding,
and we intend for this resource to evolve as new results become available, and relevant
existing models are added to \stdpopsim by the community.
We have implemented both a command line interface and a simple Python API
that can be used to simulate genomic data
from a choice of organism, genetic map, chromosome, and demographic history.
In this way, \stdpopsim will lower the barrier to high-quality simulation for exploratory analyses,
enable rigorous evaluation of population genetic software,
and contribute to increased reliability of population genetic inferences.

@@

%ACS: the consortium really isn't the point; let's get right to describing what stdpopsim is
%The first contribution of the PopSim consortium is 
The \stdpopsim library is a
a community-maintained collection of empirical genome data and population genetics simulation models,
with a structure depicted in Figure \ref{fig:cartoon}.
The package centers on a catalog of genomic information and demographic models
for a growing list of species (Fig.~\ref{fig:cartoon}A), 
and software resources to facilitate efficient simulations (Fig.~\ref{fig:cartoon}B-C).

Given the genome data and simulation model descriptions defined within the
library, it is straightforward to run standardized simulations
across a range of organisms. \Stdpopsim has a Python API and a user-friendly
command line interface, allowing users with minimal experience direct access to
state-of-the-art simulations. Simulations are output in the ``tree sequence''
format \citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring}, which
contains complete genealogical information about the simulated samples, is
extremely compact, and can be processed efficiently using the \tskit library
\citep{kelleher2016efficient,kelleher2018efficient}.
The tree sequence format could also be converted
to other formats (e.g., VCF) by the user if desired.

@@

Currently, \stdpopsim uses the  \texttt{msprime} coalescent simulator \citep{kelleher2016efficient}
as the default simulation engine.
Coalescent simulations, while highly efficient, are limited in their ability to model
continuous geography or 
complex selection scenarios, such as recurrent sweeps and background selection.
For these reasons,
we have also implemented the forward-time simulator, \texttt{SLiM} \citep{haller2019tree,haller2019slim},
as an alternative backend engine to \texttt{stdpopsim}, allowing for the simulation of processes that cannot be modeled under
the coalescent. However, as forward-time simulators explicitly model all individuals in a population, simulating large
population sizes can be highly demanding of computational resources. One common practice used to address this challenge is to
simulate a \emph{smaller} population, but to rescale resulting times,
mutation rates, recombination rates, and selection coefficients
so that the intensity of mutation, recombination, and allele frequency change due to selection
per unit time remains the same
%% PLR: the following is not worded quite right, or at least confusingly, because if you scale time
%% then this scales rates also; I've opted for a less precise but more descriptive formulation?
% scale down the population size used in the simulations (\textit{N}),
% scale down times specified in the demographic model by the same factor, and scale up the mutation and recombination
% rates and selection coefficients such that the scaled parameters, \textit{N$\mu$}, \textit{Nr}, and \textit{Ns}, remain unchanged
\citep[see the \texttt{SLiM} manual and][]{uricchio2014robust}.
Our implementation of the \texttt{SLiM} backend allows easy use of this \emph{rescaling}
through a single ``scaling factor'' argument.
Such down-scaled simulations are not completely equivalent to simulating
all individuals in the population, and may lead to subtle differences,
especially in the presence of selection.
However, since many sequence-based measures of population diversity remain nearly unchanged when rescaling in this fashion,
this practice is effective for many purposes and widely employed.

We validated our implementation of the \texttt{SLiM} engine by comparing
estimates of several population genetic summary statistics for neutral simulations generated by both \texttt{SLiM}
and \texttt{msprime}.
Examples of this validation for the AncientEurasia\_9K19 model \citep{AncientEurasia 9K19}
are shown in Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}.
For this model,
down-scaling factors of up to $50$ produce patterns of both diversity and linkage disequilibrium that are
very similar to those observed under the coalescent (i.e., \texttt{msprime}), when both are simulated using the HapMapII\_GRCh37 recombination map
(Figure \ref{fig:slim_val_map}),
although the distribution of Tajima's $D$ differs slightly when scaling down by a factor of $50$.
%likely due to the subtle differences produced by down-scaling.
Interestingly, the effects of down-scaling are more prominent and extend to other sequence-based statistics
when data sets are simulated under a uniform recombination rate (Figure \ref{fig:slim_val_nomap}).
Importantly, both comparisons validate the equivalence of \texttt{SLiM} and \texttt{msprime} when no down-scaling is applied.
The results are also optimistic about the rescaling strategy to reduce computational burden,
but the possible effects are not well-understood,
so results relying on rescaled simulations should be carefully validated.

\subsection*{Documentation and reproducibility}

@@

% ACS: there was indeed a lot of redundancy here.  DRS improved it a lot, but I've tried to do a bit more
\subsubsection*{ Single-population demographic models.}
For single-population demographic models, we compared
\MSMC~\citep{schiffels2014inferring}, \smcpp~\citep{terhorst2017robust}, and
\stairwayplot~\citep{liu2015exploring}
 on simulated genomes sampled from a single population,
under several of the demographic models described above.
However, these experiments raise the question of what to use as the ``true" population sizes in the case of multi-population models with migration.
In particular, a simple single-population model that is fit to data simulated
under a multi-population model, 
is not expected to recover the actual simulated population sizes,
because of model misspecification.
Instead, we argue that the best one may expect in such a scenario is to infer a model that accurately reflects the
coalescence time distribution of the simulated model.
Under a multi-population model, the coalescence time distribution is influenced by migration between the target population and populations not analyzed in inference,
as well as by the ancestral effective population sizes.
The inverse coalescence rate is commonly interpreted as the effective population size,
since these are equal in a single-population model with random mating.
We thus analytically computed inverse coalescence
rates in \texttt{msprime} for each simulated model, and used them as benchmarks for the ``true'' effective population sizes.
See the Appendix for a precise definition and description of the inverse coalescence rate computation.

@@

In Figure \ref{fig:IM_popn_human}C we show estimates of population sizes and divergence
time, for each of the inference methods, using samples drawn from African and European populations
simulated under the OutOfAfrica\_3G09 model. Our results highlight many
of the strengths and weaknesses of the different methods.
For instance, the SFS-based approaches with simple IM models do not capture
recent exponential growth in the CEU population, but do consistently recover the
simulated YRI population size history. Moreover, these approaches allow 
migration rates to be estimated (Figure \ref{fig:homsap_mig_rates}), and lead to more accurate inferences
of divergence times. However, these migration rate estimates are somewhat biased.
In contrast, \smcpp is much better at capturing the recent exponential
growth in the CEU population, though it consistently underestimates divergence times
because it assumes no migration between populations (Figure \ref{fig:IM_popn_human}C).

@@

\paragraph{Benchmarking population size inference.}
We have illustrated in this paper how \stdpopsim can be used for direct
comparisons of inferential methods on a common set of simulations. Our
benchmarking comparisons have been limited, but nevertheless
reveal some informative features.
For example, at the task of estimating population size histories for simulated human
populations, we find that the sequence-based methods (\MSMC and \smcpp)
perform somewhat better overall---at least for moderate times in the past---than
the site frequency spectrum-based method (\stairwayplot),
which tends to over-estimate the sizes of oscillations
(Figures~\ref{fig:n_t_ragsdale} and~\ref{fig:n_t_gutenkunst}).
In contrast, \stairwayplot outperforms the sequence-based methods
on simulations of \textit{D.~melanogaster} or \textit{A.~thaliana} populations,
in which linkage disequilibrium is reduced (Figures~\ref{fig:n_t_sheehan} \& \ref{fig:n_t_AraTha}).
In simulations of two human populations
(Figure~\ref{fig:IM_popn_human}), \dadi and \fastsimcoal do reasonably well at
reconstructing the simulated YRI history and estimating divergence times, but struggle with the more complex simulated CEU
history, in large part because the methods assume constant population sizes.
On the other hand, \smcpp does not have the same
restrictions on its inferred history, and as a result does much better
with the CEU history but tends to underestimate divergence times
due to the assumption of no migration.
The results for the two-population \textit{D.~melanogaster} model (Figure~\ref{fig:two_popn_fly}) are generally similar.
In these comparisons, \fastsimcoal and \dadi perform almost identically, which is expected because they fit the same models to the same summaries of the data, 
differing only in how they calculate model expectations and optimize parameters.
